Towards Practical First-Order Model Counting

### 2025-07-03 doc/talk/talk.tex
Explaining the workflow

\usetikzlibrary{arrows.meta}
\usetikzlibrary{backgrounds}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{fit}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.misc}

@@

\beamertemplatenavigationsymbolsempty
\usetheme{default}
\usecolortheme{beaver}

@@

\begin{frame}{First-Order Model Counting: The Motivation}
  \begin{exampleblock}{Example Setting}
    \begin{itemize}
      \item Let \structure{$\Delta$} be a set of cardinality \structure{$n$}
      \item Suppose we want to count all \structure{$P \subseteq \Delta^{2}$}
            that are:
            \begin{itemize}
              \item functions,
              \item bijections,
              \item partial orders,
              \item symmetric,
              \item transitive,
              \item etc.
            \end{itemize}
    \end{itemize}
  \end{exampleblock}
  \pause
  \begin{itemize}
    \item[\faThumbsODown] Propositional model counting (\structure{$\#\SAT$}) is
          \alert{$\#\P$-complete}
    \item[\faThumbsOUp] But many of these counting problems have
          \alert{efficient solutions}
    \item And we can find them using \alert{first-order model counting}
          \begin{itemize}
            \item i.e., reasoning about sets, subsets, and arbitrary elements
                  without \alert{grounding} them
          \end{itemize}
  \end{itemize}
\end{frame}

% TODO:
% 1) define FOMC: any number of domains, any number of predicates of any arity
% 2) introduce WFOMC algorithms
% 3) Introduce Greedy vs BFS?
\begin{frame}{First-Order Model Counting}
\end{frame}

@@

% TODO: 4 more slides explaining the solution (maybe including the theorems)

\begin{frame}{}
\end{frame}

>>>>

\usetikzlibrary{arrows.meta}
\usetikzlibrary{backgrounds}
\usetikzlibrary{cd}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{fit}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.misc}

@@

\beamertemplatenavigationsymbolsempty
\setbeamertemplate{footline}[frame number]
\usetheme{default}
\usecolortheme{beaver}

@@

\begin{frame}{Motivation}
  \begin{exampleblock}{Example Setting}
    \begin{itemize}
      \item Let \structure{$\Delta$} be a set of cardinality \structure{$n$}
      \item Suppose we want to count all \structure{$P \subseteq \Delta^{2}$}
            (as a function of \structure{$n$}) that are:
            \begin{itemize}
              \item functions,
              \item bijections,
              \item partial orders,
              \item symmetric,
              \item transitive,
              \item etc.
            \end{itemize}
    \end{itemize}
  \end{exampleblock}
  \pause
  \begin{itemize}
    \item[\faThumbsODown] Propositional model counting (\structure{$\#\SAT$}) is
          \alert{$\#\P$-complete}
    \item[\faThumbsOUp] But many of these counting problems have
          \alert{efficient solutions}
    \item And we can find them using \alert{first-order model counting}
          \begin{itemize}
            \item i.e., reasoning about sets, subsets, and arbitrary elements
                  without \alert{grounding} them
          \end{itemize}
  \end{itemize}
\end{frame}

% TODO: (maybe later, once I know more)
% 1) define FOMC: any number of domains, any number of predicates of any arity (do I need to define weights?)
% 2) introduce WFOMC algorithms
% 3) Introduce Greedy vs BFS?
\begin{frame}{First-Order Model Counting}
\end{frame}

@@

% TODO: 4 more slides explaining the solution (maybe including the theorems)
% 0. Preprocess
% 1. Use Crane to compile into functions
% 2. Simplify
% 3. FindBaseCases TODO
% 4. For each base case
% 4.1. For each constant (TODO?)
% 4.1.1. Propagate TODO
% 5. Compile to C++

\begin{frame}{Knowledge Compilation Workflow (1/2)}
  \begin{enumerate}
    \item Use \textsc{Crane} to compile \structure{$\phi$} into a set of
          equations \structure{$\mathcal{E}$}
    \item Simplify them, e.g.,
          \[
            g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
          \]
          becomes
          \[
            g(l, m) = g(l-1, m) + mg(l-1, m-1)
          \]
    \item Identify a sufficient set of base cases
          \begin{itemize}
            \item e.g., \structure{$\{\, g(0, m), g(l, 0) \,\}$}
          \end{itemize}
  \end{enumerate}
\end{frame}

% TODO:
% 2) maybe space out
% 3) add labels for each step
% 4) add question marks that then change to 0^m and 1
% 5) make arrows gray
% 6) make the last line (except C++ code) appear last
% 8) mark which steps will be explained in more detail
\begin{frame}[fragile]{Knowledge Compilation Workflow (2/2)}
  \centering For each base case:

  \begin{tikzcd}
    g(0, m) \arrow[d] & \phantom{-} \arrow[d, phantom] & g(l, 0) \arrow[d] \\
    \forall x \in \Gamma\text{. }S(x) \lor \neg S(x) \arrow[d] & \phantom{-} & \top \arrow[d] \\
    g(0, m) = 0^m \arrow[rd] & & g(l, 0) = 1 \arrow[ld] \\
    & \mathcal{E} \arrow[d] & \\
    \text{Domain sizes} \arrow[r] & \text{C++ code} \arrow[r] & \text{Model count}
  \end{tikzcd}
\end{frame}
